---
title: "Lab 03 - Classification"
author: "Ken Ye"
date: "9/13/2023"
output:
  html_document:
    df_print: paged
---

## 1. Stock Market Data

This lab will work with the `Smarket` stock market data from the `ISLR` library.
The data consists of the percentage returns for the S&P 500 stock index for each
day between the beginning of 2001 to the end of 2005. Each date has information
about the percentage returns for the 5 previous trading days (`Lag1` through
`Lag5`), the `Volume` (number of shares traded on the previous day), `Today`
(percentage return on given date) and `Direction` (if the market was up or down
on this date.)

We'll start by loading and examining the data.

```{r}
library(ISLR)
attach(Smarket)
names(Smarket)
pairs(Smarket)

```

The `cor()` function can be used to calculate the matrix of all pairwise
correlations between predictors.

```{r}
cor(Smarket[, -9])
```

We can also look at the `Volume` variable over the time period considered.

```{r}
plot(Volume)
```

## 2. Logistic Regression

The variable of interest is predicting the `Direction`, whether a stock moved
`Up` or `Down`. We will fit a logistic regression model to predict `Direction`
using the other variables (except `Today`).

The `glm()` function can be used to fit generalized linear models, of which
logistic regression is an example. The functionality of `glm()` is similar to
that of the `lm()` function from last week. We will need to pass the argument
`family = binomial` to the `glm()` function to specify logistic regression.

```{r}
glm.fit <- glm(Direction ~ Lag1 + Lag2 + Lag3 + Lag4 + Lag5 + Volume, 
               data = Smarket, family = binomial)
summary(glm.fit)
```

The `summary()` and `coef()` functions can also be used with `glm()` model fits.

```{r}
coef(glm.fit)
summary(glm.fit)$coef
summary(glm.fit)$coef[, 4]
```

The `predict()` function can also be used, similar to its use with the `lm()`
function.

```{r}
glm.probs <- predict(glm.fit, type = "response")
glm.probs[1:10]
```

The `contrasts()` function can be used to see the dummy variable encoding for
the `Direction` variable.

```{r}
contrasts(Direction)
```

This means that when `Up = 0`, `Direction = Down`, while when `Up=1`, we have
that `Direction = Up`.

Next, we will want to convert the predicted probabilities to either `Up` or
`Down` based on the value of the probability. Probabilities greater than 0.5
will be classified as `Up`.

```{r}
glm.pred <- rep("Down", 1250)
glm.pred[glm.probs > 0.5] <- "Up"
```

We can create a confusion matrix using the `table()` function. The confusion
matrix gives us the number of True Positives, True Negatives, False Positives
and False Negatives.

```{r}
table(glm.pred, Direction)
```

We can calculate the accuracy of the classifier by finding the percentage of
points that are correctly classified.

```{r}
mean(glm.pred == Direction)
```

Now, we will divide our dataset into training and validation sets. Observations
from 2001-2004 will be in the training set and observations from 2005 will be in
the validation set.

```{r}
train <- (Year < 2005)
Smarket.2005= Smarket [! train ,]
dim(Smarket.2005)
Direction.2005=Direction[!train]
```

We can again fit the logistic regression model using `glm()`, this time
restricted to just our training set. We can then predict the `Direction` on the
validation set using the `predict()` function.

```{r}
glm.fit <- glm(Direction ~ Lag1 + Lag2 + Lag3 + Lag4 + Lag5 + Volume, 
               data = Smarket, family = binomial, subset = train)
glm.probs <- predict(glm.fit, Smarket.2005, type = "response")
```

Then, we can compare the predictions on the validation set to the true labels
and calculate the accuracy of our logistic regression classifier.

```{r}
glm.pred <- rep("Down", 252)
glm.pred[glm.probs > 0.5] <- "Up"
table(glm.pred, Direction.2005)
mean(glm.pred == Direction.2005) ## Accuracy
mean(glm.pred != Direction.2005) ## Misclassification rate
```

We can improve the predictive performance of our classifier by considering a
subset of the predictor variables; only those that have the strongest
relationship to the response. From the correlation matrix calculated above, we
can see that these variables are `Lag1` and `Lag2`.

```{r}
glm.fit <- glm(Direction ~ Lag1 + Lag2, data = Smarket, family = binomial, subset = train)
glm.probs <- predict(glm.fit, Smarket.2005, type = "response")
glm.pred <- rep("Down", 252)
glm.pred[glm.probs > 0.5] <- "Up"
table(glm.pred, Direction.2005)
```

The predictive performance has slightly improved by limiting the predictor
variables to a subset.

```{r}
mean(glm.pred == Direction.2005)

```

We can also predict the `Direction` for two new days with the lags given below
using the `predict()` function.

```{r}
predict(glm.fit,newdata=data.frame(Lag1=c(1.2,1.5),
                                   Lag2=c(1.1,-0.8)),type="response")
```

## 3. K-Nearest Neighbors

To fit a K-nearest neighbors model, we need to use the `class()` package. First,
however, we need to split our data into training and validation sets. We can use
`cbind()` (column bind) to bind `Lag1` and `Lag2` into a matrix for each subset.

```{r}
library(class)
train.X <- cbind(Lag1, Lag2)[train, ]
test.X <- cbind(Lag1, Lag2)[!train, ]
train.Direction <- Direction[train]
```

We need to use `set.seed()` to initialize the random number generator for
consistent results and then can call the `knn()` function to fit the classifier
and make predictions about `Direction`.

```{r}
set.seed(1)
knn.pred <- knn(train.X, test.X, train.Direction, k = 1)
table(knn.pred, Direction.2005)
```

The accuracy of the classifier on the test set can be found using the diagonal
of the confusion matrix.

```{r}
accuracy <- sum(diag(table(knn.pred, Direction.2005)))/nrow(Smarket)
```

We can also repeat the fit with $K=3$ and compute the confusion matrix and
accuracy.

```{r}
knn.pred <- knn(train.X, test.X, train.Direction, k = 3)
table(knn.pred, Direction.2005)
mean(knn.pred == Direction.2005)
```

## 4. Application

We can also apply the KNN classifier to the `Caravan` dataset. The response
variable is `Purchase`, which indicates whether or not an individual purchases a
caravan insurance policy. This dataset is an imbalanced classification problem,
as only 6% of people in the dataset purchase the caravan insurance.

```{r}
attach(Caravan)
dim(Caravan)
summary(Purchase)
348/5822
```

We can use the `scale()` function to scale the data to have a mean of 0 and a
standard deviation of 1.

```{r}
standardized.X <- scale(Caravan[, -86])
var(Caravan[, 1])
var(Caravan[, 2])
var(standardized.X[,1])
var(standardized.X[,2])
```

Again, we can split the data into a training and test set and make predictions
about the response variable `Purchase` using a KNN model.

```{r}
test <- 1:1000
train.X <- standardized.X[-test, ]
test.X <- standardized.X[test, ]
train.Y <- Purchase[-test]
test.Y <- Purchase[test]
set.seed(1)
knn.pred <- knn(train.X, test.X, train.Y, k = 1)
mean(test.Y != knn.pred)

mean(test.Y != "No")

table(knn.pred, test.Y)

9/(68 + 9)


```

We can also fit the model with other values of $K$, for example $K=3$ and $K=5$.

```{r}
knn.pred <- knn(train.X, test.X, train.Y, k = 3)
table(knn.pred, test.Y)
5/26
knn.pred <- knn(train.X, test.X, train.Y, k = 5)
table(knn.pred, test.Y)

4/15
```

Finally, we can compare the KNN model with a logistic regression model fit using
`glm()`.

```{r}
glm.fit <- glm(Purchase ~ ., data = Caravan, family = binomial, subset = -test)
glm.probs <- predict(glm.fit, Caravan[test, ], type = "response")
glm.pred <- rep("No", 1000)
glm.pred[glm.probs > 0.5] <- "Yes"
table(glm.pred, test.Y)

glm.pred <- rep("No", 1000)
glm.pred[glm.probs > 0.25] <- " Yes"
table(glm.pred, test.Y)

11/(22 + 11)

```

## 5. Exercises

### 1. Logistic Regression

We will again use the `iris` data for classification.

```{r}
data("iris")
colnames(iris)
attach(iris)
```

(a) Find the number of unique `Species` for the `iris` data.

```{r}

```

(b) Since there are more than 2 classes for the `iris` data, we will consider a
    1-vs-rest approach to classification. Add a new column to the `iris` dataset
    that has the labels that we wil use for classificaition. Call this column
    `Species2`. Encode this column such that the `setosa` is the positive class
    (`Speicies2 = 1`), while the other Species correspond to the negative class
    (the outcome variable is 0 for these other classes).

```{r}

```

(c) Next, we need to divide our data into a training and validation set. Examine
    the `iris` data and propose a method for this division. What do we need to
    be careful about based on the original `iris` data?

```{r}
## Need to shuffle data, all classes in order right now

```

(d) What are some cases where we need to be careful how we split our data into
    training and validation sets? What is the usual assumption when we split the
    data randomly?

(e) Now, split the `iris` data into a training and validation set. Use 30% of
    the data for the validation set. Make sure to use `set.seed()` (see Lab 01
    for a reminder about this function). Print the dimensions of you training
    and validation sets.

```{r}

```

(f) Fit a logistic regression classifier to your `iris` training data. Display a
    summary of the model fit. Use `Sepal.Length`, `Sepal.Width`, `Petal.Length`
    and `Petal.Width` as predictors.

```{r}

```

(g) Explore adding some interaction terms or dropping terms from the model that
    do not appear significant. You do not need to perform forward/backward
    selection, just explore the data and model fits. Interpret the parameters in
    your final model.

```{r}

```

(h) Use the `predict()` function to predict the species for your validation
    data. Use `predict()` for both models, from part (f) and part (g).

```{r}

```

(i) Calculate the confusion matrix and accuracy for your two models, from parts
    (f) and (g). Make sure to first convert your predicted probabilities from
    part (h) to `Species2`. Assume that a probability of greater than 0.5
    corresponds to the positive class species, i.e. `Species2=1` and
    `Species = setosa`. Which model would you select and why?

```{r}

```

### 2. Conceptual

Think about the `Caravan` example above. Is accuracy a good measure to evaluate
potential classifiers on this data set? Why or why not? What might be some
better metrics to use to evaluate classifiers for imbalanced data sets?
