---
title: "HW3"
author: "Ken Ye"
date: "`r Sys.Date()`"
output: pdf_document
---

# Question 4 (ISL Chapter 7, Exercise 9)

This question uses the variables dis (the weighted mean of distancesto five
Boston employment centers) and nox (nitrogen oxides concentrationin parts per 10
million) from the Boston data. We will treatdis as the predictor and nox as the
response.

a\. Use the poly() function to fit a cubic polynomial regression to predict nox
using dis. Report the regression output, and plot the resulting data and
polynomial fits.

```{r}

```

(b) Plot the polynomial fits for a range of different polynomialdegrees (say,
    from 1 to 10), and report the associated residualsum of squares.

    ```{r}

    ```

(c) Perform cross-validation or another approach to select the optimaldegree for
    the polynomial, and explain your results.

(d) Use the bs() function to fit a regression spline to predict noxusing dis.
    Report the output for the fit using four degrees offreedom. How did you
    choose the knots? Plot the resulting fit.

(e) Now fit a regression spline for a range of degrees of freedom, andplot the
    resulting fits and report the resulting RSS. Describe theresults obtained.

(f) Perform cross-validation or another approach in order to selectthe best
    degrees of freedom for a regression spline on this data.Describe your
    results.
